---
title: "Exploring FIFA Ratings and motivation"
output: html_document
date: "2024-01-25"
---



```{r setup, include=FALSE}
library(tidyverse)
library(lmtest)
library(sandwich)
library(lfe)
library(MASS)
library(glmnet)
library(pROC)
library(ordinalNet)
library(orf)
library(ggplot2)
library(car)
rm = list(ls())
```

``` {r loading the data, include=FALSE}

data <- read_csv('train.csv')
oos_data <- read_csv("test.csv")
```


```{r plotting density plots}

ggplot(data=data, aes(x=HomePlayer_Overall_mean)) +
  geom_density(fill="tan1") + labs(title="OVR Distribution") + theme(plot.title = element_text(hjust = 0.5))

ggplot(data=data, aes(x=HomePlayer_Age_mean)) +
  geom_density(fill="tan1") + labs(title="Age Distribution") + theme(plot.title = element_text(hjust = 0.5))

ggplot(data=data, aes(x=HomePlayer_Overall_sd)) +
  geom_density(fill="tan1") + labs(title="Overall SD distribution") + theme(plot.title = element_text(hjust = 0.5))

ggplot(data=data, aes(x=HomePlayer_Weight_mean)) +
  geom_density(fill="tan1") + labs(title="Weight Distribution") + theme(plot.title = element_text(hjust = 0.5))

ggplot(data=data, aes(x=HomePlayer_Height_mean)) +
  geom_density(fill="tan1") + labs(title="Height Distribution") + theme(plot.title = element_text(hjust = 0.5))



```


```{r}
ggplot(data = data, aes(x = HomePlayer_Overall_mean,
                       y = home_score)) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home Overall and the home_score")
ggplot(data = data, aes(x = HomePlayer_Weight_mean,
                       y = home_score)) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home Weight and the home_score")
ggplot(data = data, aes(x = HomePlayer_Height_mean,
                       y = home_score)) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home Height and the home_score")
ggplot(data = data, aes(x = HomePlayer_Age_mean,
                       y = home_score)) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home Age and the home_score")

ggplot(data = data, aes(x = home_GD_prior- away_GD_prior,
                       y = home_score)) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home Age and the home_score")

ggplot(data = data, aes(x = home_Points_form_pw,
                       y = log(home_score+1))) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home Age and the home_score")

ggplot(data = data, aes(x = home_points_to_rel,
                       y = away_points_to_rel)) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home Age and the home_score")
```

Now we will see different models to check the following:

1. Does the standard deviation have a positive effect on the teams performance? 
2. How is the motivation to not get relegated affects the team different than the motivation to reach top 4 / win the league
3. Which is better in estimating form? Goal differences before the game or points gained

First we will see which is better at estimating form: points form or GD form over the past 5-6 (per week)
```{r first seeing a model only with standard deviation with form}
data <- data%>%
  mutate(GD = home_score-away_score)
GD_model <- felm(data=data, formula = GD ~
                HomePlayer_Overall_mean+
              HomePlayer_Overall_sd +
                AwayPlayer_Overall_mean+
                AwayPlayer_Overall_sd+
                home_GD_form_pw + 
                away_GD_form_pw 
               | home_team_name + away_team_name | 0 | home_team_name)

Points_model <- felm(data=data, formula = GD ~
                HomePlayer_Overall_mean+
              HomePlayer_Overall_sd +
                AwayPlayer_Overall_mean+
                AwayPlayer_Overall_sd+
                home_Points_form_pw + 
                away_Points_form_pw 
               | home_team_name + away_team_name | 0 | home_team_name)

summary(GD_model)
summary(Points_model)
```
Now we see looking at the adjusted R squared for the full model, it's better off to use to GD which makes sense because there will be instances where 2 teams are in the same form point wise but one team scores more goals.

Now we will add motivation to the mix:

```{r adding motivation}
staying_league_model <- felm(data=data, formula = GD ~
                HomePlayer_Overall_mean + AwayPlayer_Overall_mean+
              HomePlayer_Overall_sd+
                AwayPlayer_Overall_sd+
                +home_GD_form_pw + away_GD_form_pw+
                home_points_to_rel + away_points_to_rel
               | home_team_name + away_team_name | 0 |home_team_name)


tr_model <- felm(data=data, formula = GD ~
                HomePlayer_Overall_mean + AwayPlayer_Overall_mean+
              HomePlayer_Overall_sd+
                AwayPlayer_Overall_sd+
                 +home_GD_form_pw + away_GD_form_pw+
                home_points_to_championship + away_points_to_championship
               | home_team_name + away_team_name | 0 | home_team_name)

gi_model <- felm(data=data, formula = GD ~
                HomePlayer_Overall_mean + AwayPlayer_Overall_mean+
              HomePlayer_Overall_sd+
                AwayPlayer_Overall_sd+
                 +home_GD_form_pw + away_GD_form_pw+
                home_match_importance + away_match_importance
               | home_team_name + away_team_name | 0 | home_team_name)

summary(gi_model)
summary(tr_model)
summary(staying_league_model)

linearHypothesis(gi_model, c("home_match_importance= 0", "away_match_importance=0"))
linearHypothesis(tr_model, c("home_points_to_championship= 0", "away_points_to_championship=0"))
linearHypothesis(staying_league_model, c("home_points_to_rel= 0", "away_points_to_rel=0"))

ggplot(data = data, aes(x = HomePlayer_Overall_mean,
                       y = home_points_to_rel)) +
  geom_point() + stat_smooth(method = "lm") +
  ggtitle("Home OVR mean and match importance")
```

Naively we can say there is a problem when motivation, but that can be due to the high correlations between both the teams quality, and their position in the league, and we can gain information mainley from non linear models that can use such information to predict better.
